# Results

## Question 1: How big of a problem actually is Long COVID? Have reported cases increased over the past year, decreased, or remained relatively constant, and what percentage of adults in the US are actually impacted?

Step 1: need to load in the data and process it
```{r}
# Import Required Libraries
library(dplyr)
library(ggplot2)
library(forcats)

# Import Data
pcc <- read.csv("Post-COVID_Conditions.csv")
```

Step 2: Select only the data we want. For this question, since we are just looking at the national impact of Long COVID, we will only select the rows contianing a value of "National Estimate" in the `Group` column. Additionally, we only want the columns for `Indicator`, `Group`, `Time.Period.Start.Date`, `Time.Period.End.Data`, and `Value`.

```{r}
TimeData <- pcc %>% filter(Group == 'National Estimate') %>% select(c(Indicator, Group, Time.Period.Start.Date, Time.Period.End.Date, Value))

# Also convert the Time Period Colums to dates
TimeData$Time.Period.Start.Date <- as.Date(TimeData$Time.Period.Start.Date, format = "%m/%d/%Y")
#TimeData$Time.Period.Start.Date <- format(TimeData$Time.Period.Start.Date, "%B %d, %Y")

TimeData$Time.Period.End.Date<- as.Date(TimeData$Time.Period.End.Date, format = "%m/%d/%Y")
#TimeData$Time.Period.End.Date <- format(TimeData$Time.Period.End.Date, "%B %d, %Y")
```

With the data filtered, we can now look at each indicator one at a time to identify overall trends in the United States. To test the code, we're going to first try this with national estimates for anyone that every had COVID, stored in rows with `Indicator` = "Currently experiencing long COVID" (either as a percentage of adults who ever had COVID, or as a percentage of all adults).

```{r}
CurrentExp <- TimeData %>% filter(Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of all adults")

CurrentExp <- na.omit(CurrentExp)

CurrentExp$Indicator <- recode(CurrentExp$Indicator, "Currently experiencing long COVID, as a percentage of adults who ever had COVID" = "As a % of Adults who have ever had COVID", "Currently experiencing long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(CurrentExp, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently Experiencing Long COVID") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "2 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```

From the graph, we see that overall, the proportion of adults that are currently experiencing Long COVID has consistently decreasd over the course of the data collection across both cohorts, particularly when looking at the adults that are currently experiencing Long COVID and comparing it to all adults who have ever had COVID.

Now, we want to look at the percentage of adults who have EVER experienced Long COVID.

```{r}
EverExp <- TimeData %>% filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Ever experienced long COVID, as a percentage of all adults")

EverExp <- na.omit(EverExp)

EverExp$Indicator <- recode(EverExp$Indicator, "Ever experienced long COVID, as a percentage of adults who ever had COVID" = "As a % of Adults who have ever had COVID", "Ever experienced long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(EverExp, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Ever Experienced Long COVID") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "2 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```

This seems like an error, or at least something funky with the data collection, as the total should not decrease.

## Question 3: What impact does Long COVID have on various health metrics, such as level of fatigue or activity metrics?

In our data set, the CDC measured the impact of Long COVID on overall health using questions about activity, looking at both those currently experiencing any activity limitations due to Long COVID and those currently experiencing significant activity limitations due to Long COVID (for more details as to the differences between these groups, please see the data overview section). 

To get an overview of the effect of Long COVID on health, we first want to see if Long COVID truly has a tangible impact on health. To do so, we used national estimates of Long COVID (stored in rows contianing a value of "National Estimate" in the `Group` column), and specifically focused on the indicators of activity limiations over time.

```{r}
# Can re-use the TimeData dataframe that we created earlier to analyze time series data for question 1:
AnyActivityLimits <- TimeData %>% filter(Indicator == "Any activity limitations from long COVID, as a percentage of adults who currently have long COVID" | Indicator == "Any activity limitations from long COVID, as a percentage of all adults")

AnyActivityLimits <- na.omit(AnyActivityLimits)

AnyActivityLimits$Indicator <- recode(AnyActivityLimits$Indicator, "Any activity limitations from long COVID, as a percentage of adults who currently have long COVID" = "As a % of Adults who have ever had Long COVID", "Any activity limitations from long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(AnyActivityLimits, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently experiencing any Activity Limitations") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "3 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```

Suggests that most adults with Long COVID are also experiencing some degree of activity limitations. Additionally, this data suggests that this problem is pervasive, as the proportion of adults experiencing Long COVID-attributed activity limitations has not decreased over the past year.

We now want to look at those experiencing **significant** activity limitations, in a similar analysis to the above.

```{r}
SignifActivityLimits <- TimeData %>% filter(Indicator == "Significant activity limitations from long COVID, as a percentage of adults who currently have long COVID" | Indicator == "Significant activity limitations from long COVID, as a percentage of all adults")

SignifActivityLimits <- na.omit(SignifActivityLimits)

SignifActivityLimits$Indicator <- recode(SignifActivityLimits$Indicator, "Significant activity limitations from long COVID, as a percentage of adults who currently have long COVID" = "As a % of Adults who have ever had Long COVID", "Significant activity limitations from long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(SignifActivityLimits, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently experiencing Significant Activity Limitations") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "3 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```

Same trends here are visible, albeit to a lesser extent. Suggests that of those that experience post-COVID symptoms, many experience some form of lasting, significant impacts to their overall health (aka, Long COVID is an actual problem for those afflicted).

Here, we have the option to add more about if specific groups are more affected than others. Check back on this after meeting.
