# Results

Using the graphs below, we attempted to get an understanding regarding which factors seem to have an effect on the likelihood of developing long COVID and how large of an effect that can be.

```{r}
# Preprocessing

pcc <- pcc %>%
  mutate(Phase = factor(ifelse(Phase == 3.1 ,"3.10", as.character(Phase)), levels = c("3.5","3.6","3.7","3.8","3.9","3.10"))) %>%
  mutate(Subgroup = factor(Subgroup,levels(fct_relevel(pcc$Subgroup,"Less than a high school diploma","High school diploma or GED","Some college/Associate's degree","Bachelor's degree or higher")))) %>%
  select(Indicator, Group, Subgroup, Phase, Value) %>%
  na.omit()
```

```{r}
pcc %>%
  filter(Group == "By Age") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Age Group", x = "Age Group", y = "Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  
  coord_flip()
```

From the graph we can observe that age seems to play a role, to some extend, since for all phases the same pattern was observed. Middle aged people seem to be more susceptible, with a peak at 50-59 years old.

However, as time passes, the differences seem to get smaller between age groups, almost disappearing but the pattern for the people who claim that they have experienced long COVID at some point, remains unchanged. It is also worth pointing out that the percentage of people above 80 years old experiencing long COVID during phases 3.6 and 3.8 increased.

Lastly, we can notice that while the percentages of people below the age of 40 experiencing long COVID during each period are lower than that of people above the age of 60, the opposite can be observed when it comes to the percentages of people ever experiencing long COVID. This could be an indicator that younger people might be more susceptible to experiencing long COVID but their symptoms pass quicker compared to older people who might be less susceptible but experience the symptoms for longer.

```{r}
pcc %>%
  filter(Group == "By Sex") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Sex", x = "Sex", y = "Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```

Regarding Sex as a factor, it seems to play a more crucial role since the differences between the percentages of females and males are much more significant. Females have been and still are more susceptible than men , more specifically men's percentages are 2/3 of the female ones.

```{r}
pcc %>%
  filter(Group == "By Race/Hispanic ethnicity") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Race/Hispanic Ethnicity", x = "Race/Hispanic Ethnicity", y = "Average Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```

Concerning the Race/Hispanic Ethnicity, Non-Hispanic Asian people are by far the least susceptible to developing long COVID while the Non-Hispanic of multiple races or races other than White, Black, Asian or Hispanic/Latino are the most susceptible. All the other races stand. in between and have almost equal percentages.

```{r}
pcc %>%
  filter(Group == "By Gender identity") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Gender Identity", x = "Gender Identity", y = "Average Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
    coord_flip()
```

We already covered the comparison between males and females but now we can notice that transgender individuals' percentages are 2-3 times larger than those of males and sometimes double of that of females as well. Those differences are observed among the persons experiencing long COVID each period but for the ones claiming to ever have experienced, the pattern is the same but the differences are not that big. The percentages of transgender people still remain almost double of the corresponding male ones.

However, we can notice that the dataset is not so trustworthy because of the sampling not being representative. one would expect the "ever experienced" percentage to rise during every period since this is a cumulative indicator by default. In reality though, the percentages of the cis-gender individuals remain relatively stable while for the transgender ones there is a significant increase on phases 3.5 and 3.8.

The fact that the percentages of each category remain stable can maybe be explained by the fact that the long COVID symptoms remain for years sometimes so since the data were gathered bi-weekly, each phase lasts 2 months and we don't have more than 1.5 years of data, not enough time passed for others to pinpoint their symptoms as long COVID ones.

```{r}
pcc %>%
  filter(Group == "By Sexual orientation") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Sexual orientation", x = "Sexual orientation", y = "Average Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```

We wouldn't expect the sexual orientation to play a role to whether someone develops long COVID but according to those graphs we can deduct that while the percentages of straight and gay/lesbian are roughly equal, the percentages of bisexual people are about one third higher for almost all periods.

```{r}
pcc %>%
  filter(Group == "By Education") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "identity", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Education", x = "Education level", y = "Average Percentage") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```

When the level of education is concerned, we can observe that the higher the educational level, the lower the percentages were at first. But as the time/phases pass, the percentage of people having a bachelor's degree or higher lies always a little lower while all the other percentages are roughly equal, with the percentage of people with some college/associate's degree always being slighlty higher.

```{r}
pcc %>%
  filter(Group == "By Disability status") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently experiencing long COVID", "Ever experienced long COVID")) +
  labs(title = "Average Percentage per Disability status", x = "Disability status", y = "Average Percentage")+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```

Whether someone is disabled or not seems to be the most impactful factor, since disabled people percentages are 2-3 times higher than those of the non-disabled ones and this pattern can be observed during all phases and both among those claiming to be currently experiencing during each phase and those claiming to have ever experienced.
