# Results

## Investigating the Trends in Long COVID Prevalence

Our initial objective was to gain insight into the prevalence of Long COVID in the United States. Ultimately, our goal is to figure out what groups are affected most by Long COVID in order to best inform health policy officials when it comes to allocation of resources for diagnosis and treatment, but if Long COVID has low prevalence or shows decreasing trends in prevalence, these interventions might have limited efficacy. As such, it is important to first establish a baseline of what overall proportion of adults in the United States are either currently experiencing post-COVID-related symptoms, or have experienced such symptoms in the past.

To answer this question best, we chose to look at only rows in our dataset containing the National Estimates for Long COVID cases among adults in the U.S., as these rows will give us the best overview of trends in cases. We did so by first filtering our data to select the rows contianing a value of "National Estimate" in the `Group` column, and then only chose to keep the columns for `Indicator`, `Group`, `Time.Period.Start.Date`, `Time.Period.End.Data`, and `Value`.
```{r}
#| warning: false

# Import Required Libraries
library(dplyr)
library(ggplot2)
library(forcats)

# Import Data
pcc <- read.csv("Post-COVID_Conditions.csv")

# Filter the data according to the workflow above
TimeData <- pcc %>% filter(Group == 'National Estimate') %>% select(c(Indicator, Group, Time.Period.Start.Date, Time.Period.End.Date, Value))

# Also need to convert the Time Period Colums to dates
TimeData$Time.Period.Start.Date <- as.Date(TimeData$Time.Period.Start.Date, format = "%m/%d/%Y")
#TimeData$Time.Period.Start.Date <- format(TimeData$Time.Period.Start.Date, "%B %d, %Y")

TimeData$Time.Period.End.Date<- as.Date(TimeData$Time.Period.End.Date, format = "%m/%d/%Y")
#TimeData$Time.Period.End.Date <- format(TimeData$Time.Period.End.Date, "%B %d, %Y")
```
With the data filtered, we now looked at each indicator one at a time to identify overall trends in Long COVID cases in the United States. Since we wanted to look at the prevalence of Long COVID, we first looked at adults that are currently experiencing Long COVID, either as a percentage of adults who ever had COVID or as a percentage of all adults in the U.S. 
```{r}
CurrentExp <- TimeData %>% filter(Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of all adults")

CurrentExp <- na.omit(CurrentExp)

CurrentExp$Indicator <- recode(CurrentExp$Indicator, "Currently experiencing long COVID, as a percentage of adults who ever had COVID" = "As a % of Adults who have ever had COVID", "Currently experiencing long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(CurrentExp, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently Experiencing Long COVID") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "2 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```
From the graph, we see that overall, the proportion of adults that are currently experiencing Long COVID has consistently decreased over the course of the data collection across both cohorts, particularly when looking at the adults that are currently experiencing Long COVID and comparing it to all adults who have ever had COVID. However, this trend seems to be leveling off around 10%, suggesting that around 10% of adults that get COVID also suffer from post-COVID symptoms of some severity. Since this percentage is relatively high, support for Long COVID diagnosis and treatment still should be advised.

The dataset also contains responses for adults that have ever experienced Long COVID, either as a percentage of adults who have ever had COVID or as a percentage of all adults. In order to see how prevalent Long COVID has been over the course of the pandemic, we also wanted to see how trends in the overall number of adults impacted by Long COVID have changed over the survey duration. For this analysis, we again only selected rows in our dataset containing the National Estimates for Long COVID cases among adults in the U.S.

```{r}
EverExp <- TimeData %>% filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Ever experienced long COVID, as a percentage of all adults")

EverExp <- na.omit(EverExp)

EverExp$Indicator <- recode(EverExp$Indicator, "Ever experienced long COVID, as a percentage of adults who ever had COVID" = "As a % of Adults who have ever had COVID", "Ever experienced long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(EverExp, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Ever Experienced Long COVID") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "2 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```
In the graph above, we expect the proportion of adults who have ever experienced Long COVID to **increase**, as each time the survey is answered, all people that have previously had Long COVID should respond yes, along with all people that have newly contacted Long COVID. While this trend is seen in the proportion of adults who have ever experienced Long COVID as a percentage of **all adults**, it is not the case for the proportion of adults who have ever experienced Long COVID as a percentage of all adults who **ever had COVID**. It is possible that this result is an artifact of the data collection process, but the exact reason cannot be uncovered by information in the dataset alone.

## Determining the Key Factors that Affect Long COVID Suspectibility

We next sought to better understand what demographic factors, including age, gender, and geographic location, played a role in Long COVID susceptibility. To do so, we first broke up our data into Phases, so that we could analyze several months of data at a time (for more information on the Phases of survey information collection, see the **Data** section). This filtering required some initial processing of the data, which we performed below.
```{r}
# Load in the data with the required modifications to Phase (convert to a factor) and Subgroup (convert to a factor)
# Also only want to select the columns we want: store as a new df, RenamedPcc
RenamedPcc <- pcc %>%
  mutate(Phase = ifelse(Phase == 3.1 ,"3.10", as.character(Phase))) %>%
  mutate(Phase = recode(Phase, "3.5" = "Phase 3.5", "3.6" = "Phase 3.6", 
         "3.7" = "Phase 3.7", "3.8" = "Phase 3.8", "3.9" = "Phase 3.9", "3.10" = "Phase 3.10")) %>%
  mutate(Phase = factor(Phase, levels = c("Phase 3.5","Phase 3.6",
                                          "Phase 3.7","Phase 3.8","Phase 3.9","Phase 3.10"))) %>%
  mutate(Subgroup = factor(Subgroup,levels(fct_relevel(pcc$Subgroup, 
                                                       "Less than a high school diploma", 
                                                       "High school diploma or GED",
                                                       "Some college/Associate's degree",
                                                       "Bachelor's degree or higher")))) %>%
  select(Indicator, Group, Subgroup, Phase, Value) %>%
  na.omit()
```

With our data processing complete, we could now look at the impact of individual demographic factors on Long COVID susceptibility, measured by both the percentage of adults that are currently experiencing Long COVID and the percentage of adults that have ever experienced Long COVID. For this analysis, both these proportions are taken out of all adults who have ever had COVID, as this metric is less impacted by variation in how many people with COVID are surveyed. The first factor we looked at was **age**, which separated adults based on 10-year age ranges.
```{r}
# Generate the first series of bar charts: looking at the effect of age on the % of adults experiencing Long COVID
RenamedPcc %>%
  filter(Group == "By Age") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | 
           Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_x_discrete(labels = c("18-29 years", "30-39 years", "40-49 years", "50-59 years", 
                              "60-69 years", "70-79 years", "≥ 80 years")) +
  scale_fill_discrete(labels = c("Currently Experiencing Long COVID", "Ever Experienced Long COVID"), 
                      name = "Condition:") +
  labs(title = "% of Adults out of All Adults who Ever had COVID, separated by Age", 
       x = "Age Range", y = "Percentage") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  coord_flip()
```
From the graph above, we see that age does seem to affect likelihood to develop Long COVID, as certain age ranges have greater measured proportions of those currently experiencing Long COVID and those that ever experienced Long COVID than other age ranges. In particular, middle-aged people seem to be most susceptible, with a peak for both currently experiencing and ever experienced Long COVID at 50-59 years old across most of the different phases of survey collection. It is also interesting to note here that while the trends for "Ever Experienced Long COVID" do not change much across the various phases, the trends for "Currently Experiencing Long COVID" do. Most notably, the proportion of adults at least 80 years old experiencing Long COVID stays consistently around or above 10% during each phase, even as the proportions for the other age ranges drops. While the exact reason for this trend is unknown, one possible explanation could be that despite the lower number of 80 + year olds affected by Long COVID, those affected tend to stay affected for longer periods of time.

We next sought to analyze the effect of gender identity on Long COVID susceptibility, in a similar analysis to the one above. Again, we separated our data by survey collection phases, and took all proportions out of all adults who ever had COVID, for the same reasons as before.
```{r}
# Second bar chart set: looking at the effect of Gender Identity on the % of adults experiencing Long COVID
RenamedPcc %>%
  filter(Group == "By Gender identity") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  mutate(Subgroup = factor(Subgroup, levels = c("Cis-gender male", "Cis-gender female", "Transgender"))) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently Experiencing Long COVID", "Ever Experienced Long COVID"), 
                      name = "Condition:") +
  labs(title = "% of Adults out of All Adults who Ever had COVID,\nseparated by Gender Identity", 
       x = "Gender Identity", y = "Percentage") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", plot.margin = margin(r = 0)) +
  coord_flip()
```
From this graph, we see that cis-gender males are the least susceptible to Long COVID, while cis-gender females and people that identify as transgender were the most susceptible. In particular, transgender individuals' percentages are 2-3 times larger than those of cis-gender males, and for certain survey phases almost double of that of cis-gender females as well. However, we also noticed similar issues with the "Ever Experienced Long COVID" survey results to earlier analyses in this graph. Assuming that data collection is consistent, the values of "Ever Experienced Long COVID" should **not** decrease over time, as this is a cumulative measure. That being said, in reality, we see that cumulative proportions for all three gender identities drop across the 6 phases, which is contrary to the expected trend.

The third factor we looked at was if certain races or ethnicities were more affected by Long COVID, in a similar analysis to the previous ones.
```{r}
# Third bar chart set: looking at the effect of Race on the % of adults experiencing Long COVID
RenamedPcc %>%
  filter(Group == "By Race/Hispanic ethnicity") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  mutate(Subgroup = factor(Subgroup, levels = c("Non-Hispanic, other races and multiple races", "Hispanic or Latino",
                                                "Non-Hispanic Black, single race", "Non-Hispanic White, single race",
                                                "Non-Hispanic Asian, single race"))) %>%
  ggplot(aes(x = fct_rev(Subgroup), y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently Experiencing Long COVID", "Ever Experienced Long COVID"), 
                      name = "Condition:") +
  scale_x_discrete(labels = c("Asian", "White", "Black", "Hispanic/Latino", "Mixed/Other")) +
  labs(title = "% of Adults out of All Adults who Ever had COVID,\nseparated by Race/Ethnicity", 
       x = "Race/Ethnicity", y = "Percentage") +
  theme_linedraw() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", plot.margin = margin(r = 0)) +
  coord_flip()
```
When looking at Race/Hispanic Ethnicity, Non-Hispanic Asian people are by far the least susceptible to developing long COVID, having the lowest proportion of adults currently experiencing Long COVID and the lowest proportion of adults that have ever experienced Long COVID across all 6 phases. On the other hand, people that are either mixed race or do not identify as either Hispanic/Latino, Black, White, or Asian were the most susceptible to developing Long COVID across all 6 phases. 

Next, we looked at whether or not education status affects whether or not you experience post-COVID symptoms.
```{r}
RenamedPcc %>%
  filter(Group == "By Education") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  ggplot(aes(x = fct_rev(Subgroup), y = Value, fill = Indicator)) +
  geom_bar(stat = "identity", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently Experiencing\nLong COVID", "Ever Experienced\nLong COVID"), 
                      name = "Condition:") +
  scale_x_discrete(labels = c("Bachelor's Degree or Higher", "Some College/Associate's Degree", 
                              "High School Diploma/GED", "Less than High School Diploma")) +
  labs(title = "% of Adults out of All Adults who Ever had COVID,\nseparated by Education Level", 
       x = "Education Level", y = "Percentage") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", plot.margin = margin(r = 0)) +
  coord_flip()
```
As far as the level of education is concerned, we observe that the higher the educational level, the lower the proportion of adults affected by Long COVID. This trend is the strongest during the early phases of data collection (phases 3.5-3.7), but levels off a little as the data collection continues through phase 3.10. In fact, some violations of this trend can be seen in the later phases, with almost equal people currently experiencing Long COVID across adults without a high school diploma and adults with a bachelor's degree in phase 3.10.

We continued by looking at disability status, in particular trying to identify differences between people with and without disabilities when it comes to susceptibility to Long COVID.
```{r}
RenamedPcc %>%
  filter(Group == "By Disability status") %>%
  filter(Indicator == "Ever experienced long COVID, as a percentage of adults who ever had COVID" | Indicator == "Currently experiencing long COVID, as a percentage of adults who ever had COVID") %>%
  select(Subgroup, Phase, Value, Indicator) %>%
  mutate(Subgroup = factor(Subgroup, levels = c("Without disability", "With disability"))) %>%
  ggplot(aes(x = Subgroup, y = Value, fill = Indicator)) +
  geom_bar(stat = "summary",fun="mean", position='dodge') +
  facet_wrap(~ Phase) +
  scale_fill_discrete(labels = c("Currently Experiencing Long COVID", "Ever Experienced Long COVID"), 
                      name = "Condition:") +
  labs(title = "% of Adults out of All Adults who Ever had COVID,\nseparated by Disability Status", 
       x = "Disability Status", y = "Percentage") +
  scale_x_discrete(labels = c("Without Disability", "With Disability")) +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", plot.margin = margin(r = 0)) +
  coord_flip()
```
Whether someone is disabled or not seems to be one of the most impactful factors, since adults with disabilities have 2-3 times higher rates of Long COVID than those of the non-disabled adults. This pattern can be observed during all phases, and both among those claiming to be currently experiencing Long COVID during each phase and those claiming to have ever experienced Long COVID.

## Exploring the impact of Long COVID on activity limitations

In our data set, the CDC measured the impact of Long COVID on overall health using questions about activity, looking at both those currently experiencing any activity limitations due to Long COVID and those currently experiencing significant activity limitations due to Long COVID (for more details as to the differences between these groups, please see the data overview section). To get an overview of the effect of Long COVID on health, we now wanted to see if Long COVID truly has a tangible impact on health. To do so, we returned to using national estimates of Long COVID (stored in rows contianing a value of "National Estimate" in the `Group` column), and specifically focused on the indicators of activity limiations over time.
```{r}
# Can re-use the TimeData dataframe that we created earlier to analyze time series data for question 1:
AnyActivityLimits <- TimeData %>% filter(Indicator == "Any activity limitations from long COVID, as a percentage of adults who currently have long COVID" | Indicator == "Any activity limitations from long COVID, as a percentage of all adults")

AnyActivityLimits <- na.omit(AnyActivityLimits)

AnyActivityLimits$Indicator <- recode(AnyActivityLimits$Indicator, "Any activity limitations from long COVID, as a percentage of adults who currently have long COVID" = "As a % of Adults who have ever had Long COVID", "Any activity limitations from long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(AnyActivityLimits, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently experiencing any Activity Limitations") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "3 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```
The graph above suggests that most adults with Long COVID are also experiencing some degree of activity limitations, as throughout the past year the proportion of adults that have experienced any activity limitations out of all adults thaat have ever had Long COVID has hovered around 80%. While much lower, there is also a noticable proportion of adults with activity limitations from Long COVID out of all adults, around 6%. Additionally, this data suggests that this problem is pervasive, as the proportion of adults experiencing Long COVID-attributed activity limitations has remained consistent since September 2022.

The dataset also let us look at those experiencing **significant** activity limitations, in a similar analysis to the above. 
```{r}
SignifActivityLimits <- TimeData %>% filter(Indicator == "Significant activity limitations from long COVID, as a percentage of adults who currently have long COVID" | Indicator == "Significant activity limitations from long COVID, as a percentage of all adults")

SignifActivityLimits <- na.omit(SignifActivityLimits)

SignifActivityLimits$Indicator <- recode(SignifActivityLimits$Indicator, "Significant activity limitations from long COVID, as a percentage of adults who currently have long COVID" = "As a % of Adults who have ever had Long COVID", "Significant activity limitations from long COVID, as a percentage of all adults" = "As a % of All Adults")

ggplot(SignifActivityLimits, aes(x = Time.Period.Start.Date, y = Value, color = Indicator)) +
  geom_line() +
  geom_point() +
  ggtitle("Currently experiencing Significant Activity Limitations") + 
  xlab("Data Collection Start Date") +
  ylab("Percentage") +
  scale_x_date(labels = scales::date_format("%b, %Y"), breaks = "3 month") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(color = NULL)
```
Here, the same trends as above are visible, albeit to a lesser extent. From this information, we concluded that of those that experience post-COVID symptoms, many (around 25%) experience some form of a lasting, significant impact to their overall health. There is also a non-zero proportion of all adults in the U.S. that experience significant health ramifications related to COVID (around 1.5%), suggesting that COVID presents a significant risk for developing adverse health events even when looking at the entire U.S. population. 
